Towards Practical First-Order Model Counting

### 2024-08-09 doc/paper/paper.tex
First (second?) draft of the 'experiments' section

\begin{abstract}
  First-order model counting (FOMC) is the problem of counting the number of
  models of a sentence in first-order logic. Recently, a new algorithm for FOMC
  was proposed that, instead of simply providing the final count, generates
  definitions of (possibly recursive) functions, which can be evaluated with
  different arguments to compute the model count for any domain size. However,
  the algorithm did not include base cases in the recursive definitions. This
  work makes three contributions. First, we demonstrate how to construct
  function definitions that include base cases by modifying the logical formulas
  used in the FOMC algorithm. Second, we extend the well-known circuit
  modification technique in knowledge compilation, known as smoothing, to work
  with the formulas corresponding to base cases. Third, we introduce a
  compilation algorithm that transforms the function definitions into C++ code,
  equipped with arbitrary-precision arithmetic. These additions allow the new
  FOMC algorithm to scale to domain sizes over \num{9000} times larger than the
  current state of the art, as demonstrated through experimental results.
\end{abstract}

@@

@@

% TODO: 1. Introduction
% Objective: Briefly state the goals of the experiments.
% Overview: Summarize the structure of the section.

\begin{itemize}
  \item Note that it would not make sense to experimentally evaluate \Cranetwo{}
        without the C++ code generation described in \cref{sec:cpp} because the
        algorithm would only produce mathematical definitions of functions.
\end{itemize}

% TODO: 2. Experimental Setup

@@

\begin{itemize}
  \item The experiments were run on an Intel Skylake \SI{2.4}{\giga\hertz} CPU
        with \SI{188}{\gibi\byte} of memory and CentOS~7 operating system. C++
        programs were compiled using the Intel C++ Compiler 2020u4.
        \textsc{FastWFOMC} was run on Julia~1.10.4, while all other algorithms
        were run on the Java Virtual Machine 1.8.0\_201.
\end{itemize}

\begin{itemize}
  \item The asymptotic analysis of the total number of arithmetic operations
        required to compute the FOMC of some of these formulas was already
        performed in previous work~\cite{DBLP:conf/kr/DilkasB23}.
  \item Since FastWFOMC does not support many-sorted logic, our experiments were
        restricted to formulas with a single domain. However, many of the
        counting problems examined in [1] become equivalent (in the sense that
        they generate the same integer sequences) when restricted to a single
        domain. Moreover, it is challenging to compare Crane2 and FastWFOMC on a
        larger set of problems because each problem has to be represented in two
        (fairly different) logics: \FO{} and \UFO{}.
  \item It is important to note that comparing \Cranetwo{} and
        \textsc{FastWFOMC} on a more substantial set of benchmarks is
        challenging because there is no automated way to translate a formula in
        \FO{} or \Ctwo{} into \UFO{}, or even check if such an encoding is
        possible.
\end{itemize}

\begin{itemize}
  \item \textsc{Crane} has two modes that decide how compilation rules are
        applied to formulas: one that uses greedy search and another that
        combines greedy and breadth-first search. In our experiments, we refer
        to these modes as \Cranegreedy{} and \Cranebfs{}, respectively.
\end{itemize}

\begin{itemize}
  \item Compilation time is included in the experimental results, and the
        differences between compilation and inference time are delineated in the
        second-to-last paragraph of Section 6.
  \item We run each algorithm on each benchmark on domains of size
        $2^{0}, 2^{1}, 2^{2},\dots$ until the algorithm times out after
        \SI{1}{\hour}, runs out of memory, or (in the case of \textsc{ForcLift})
        returns $\infty$.
\end{itemize}

% TODO: 3. Results and Analysis
% Presentation of Results: Use tables, graphs, and charts to present the experimental results clearly.
% Comparison with Baselines: Compare your method’s results with the baselines.
% Discussion: Analyze and interpret the results, highlighting the strengths and potential weaknesses of your approach.
% Insights: Provide insights or patterns observed from the results.
% Challenges: Discuss any challenges or unexpected outcomes.
% Acknowledgement: Honestly acknowledge any limitations of your experiments or method.

\begin{itemize}
  \item It is important to note that while \textsc{ForcLift} may have runtime
        performance comparable to \Cranetwo{}, its finite-precision arithmetic
        cannot represent model counts for domain sizes greater than 143 and
        returns $\infty$ instead.
\end{itemize}

% TODO: 8. Conclusion
% Summary: Briefly summarize the key findings from the experimental results.
% Future Work: Suggest directions for future research based on your findings.

We compare \Cranetwo{} with \textsc{FastWFOMC} and \textsc{ForcLift} on two
problems previously considered by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. The
first problem is the function-counting problem from \cref{example:functions},
and the second problem is the bijection-counting problem described below.

The \FO{} formula for bijections is described in \cref{example:overall}. The
equivalent formula in \Ctwo{} is
\begin{align*}
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta&\text{. }P(x, y))\land{}\\
  (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta&\text{. }P(x, y)).
\end{align*}
Similarly, in \UFO{} the same formula can be written as
\begin{gather*}
  (\forall x, y \in \Delta\text{. }R(x) \lor \neg P(x, y))\land{}\\
  (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(y, x))\land{}\\
  (|P| = |\Delta|),
\end{gather*}
where $w^{-}(R) = w^{-}(S) = -1$.

@@

\Cref{fig:plot} shows the total runtime values as a measure of the overall
performance of each algorithm. We also separately tracked compilation and
inference times and commented on them where relevant. It is worth noting that
\Cranebfs{} can handle more instances than either \textsc{ForcLift} or
\Cranegreedy{}, including the bijection-counting problem in our experiments as
well as similar formulas examined previously~\cite{DBLP:conf/kr/DilkasB23}.

As shown in \cref{fig:plot}, the runtimes of all compilation-based algorithms
remained practically constant, in contrast to the rapidly increasing runtimes of
\textsc{FastWFOMC}. Although the search/compilation part is slower in
\Cranetwo{} than in \textsc{ForcLift}, the difference is negligible. The
runtimes of the FOKC algorithms appear constant because, for these counting
problems and domain sizes, compilation time dominates inference time (remember
that compilation time is independent of domain sizes). Indeed, the maximum
inference time of \Cranebfs{} and \Cranegreedy{} across these experiments is
only \SI{4}{\milli\second}. The runtimes of \Cranetwo{} have lower variation
than those of \textsc{ForcLift} because we compile the formula anew for each
domain size with \textsc{ForcLift}, whereas with \Cranetwo{}, we compile it once
and reuse the resulting C++ program for all domain sizes.

>>>>

\begin{abstract}
  First-order model counting (FOMC) is the problem of counting the number of
  models of a sentence in first-order logic. Recently, a new algorithm for FOMC
  was proposed that, instead of simply providing the final count, generates
  definitions of (possibly recursive) functions, which can be evaluated with
  different arguments to compute the model count for any domain size. However,
  the algorithm did not include base cases in the recursive definitions. This
  work makes three contributions. First, we demonstrate how to construct
  function definitions that include base cases by modifying the logical formulas
  used in the FOMC algorithm. Second, we extend the well-known circuit
  modification technique in knowledge compilation, known as smoothing, to work
  with the formulas corresponding to base cases. Third, we introduce a
  compilation algorithm that transforms the function definitions into C++ code,
  equipped with arbitrary-precision arithmetic. These additions allow the new
  FOMC algorithm to scale to domain sizes over \num{500000} times larger than
  the current state of the art, as demonstrated through experimental results.
\end{abstract}

@@

% TODO: The asymptotic analysis of the total number of arithmetic operations
% required to compute the FOMC of some of these formulas was already performed
% in previous work~\cite{DBLP:conf/kr/DilkasB23}.

@@

% 1. Introduction
% Objective: Briefly state the goals of the experiments.
% Overview: Summarize the structure of the section.

% 2. Experimental Setup

In this section, we present experimental results examining the scalability of
\Cranetwo{} compared to two other notable FOMC algorithms: \textsc{FastWFOMC}
and \textsc{ForcLift}. Note that it would not make sense to include the original
\textsc{Crane} algorithm in the experiments or to evaluate \Cranetwo{} without
the C++ code generation described in \cref{sec:cpp}. Indeed, these algorithms
can only produce mathematical definitions of functions without evaluating them.

In our experiments, we include two versions of \Cranetwo{}: \Cranegreedy{} and
\Cranebfs{}. Like its predecessor, \Cranetwo{} has two modes that decide how
compilation rules are applied to formulas: one that uses greedy search similarly
to \textsc{ForcLift} and another that combines greedy and breadth-first search.
While greedy search is more efficient (in terms of being able to handle larger
formulas), it is a heuristic that can miss some solutions.

We compare the algorithms on three benchmarks from previous work. The first is
the function-counting problem from \cref{example:functions}, previously examined
by \citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. The
second is a variant of the well-known `Friends \& Smokers' Markov logic
network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}. In \Ctwo{},
\FO{}, and \UFO{}, it can be formulated as
\begin{gather*}
  (\forall x,y \in \Delta\text{. } S(x) \land F(x, y) \Rightarrow S(y)) \land{}\\
  (\forall x \in \Delta\text{. }S(x) \Rightarrow C(x))
\end{gather*}
or, equivalently, in conjunctive normal form as
\begin{gather*}
  (\forall x,y \in \Delta\text{. }S(y) \lor \neg S(x) \lor \neg F(x, y)) \land{}\\
  (\forall x \in \Delta\text{. } C(x) \lor \neg S(x)).
\end{gather*}
Finally, we include the bijection-counting problem previously used by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. Its
formulation in \FO{} is described in \cref{example:overall}. The equivalent
formula in \Ctwo{} is
\begin{align*}
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta&\text{. }P(x, y))\land{}\\
  (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta&\text{. }P(x, y)).
\end{align*}
Similarly, in \UFO{} the same formula can be written as
\begin{gather*}
  (\forall x, y \in \Delta\text{. }R(x) \lor \neg P(x, y))\land{}\\
  (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(y, x))\land{}\\
  (|P| = |\Delta|),
\end{gather*}
where $w^{-}(R) = w^{-}(S) = -1$.

Since \textsc{FastWFOMC} does not support many-sorted logic, our experiments are
restricted to formulas with a single domain. However, many of the counting
problems examined in the experimental evaluation of
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} become equivalent (in the sense
that they generate the same integer sequences) when restricted to a single
domain. Moreover, it is challenging to compare \Cranetwo{} and
\textsc{FastWFOMC} on a larger set of problems because each problem has to be
represented in two (fairly different) logics: \FO{} and \UFO{}. Nonetheless, the
three benchmarks cover a diverse range of possibilities. Unlike the other two,
the `friends' benchmark uses multiple predicates and can be written in \FO{}
using only two variables and no cardinality constraints or counting quantifiers.
Next, the `functions' benchmark, while still liftable by all of the algorithms,
requires either the use of cardinality constraints, counting quantifiers, or
more than two variables. Finally, the `bijections' benchmark is an example of a
formula that can be handled by \textsc{FastWFOMC} but not \textsc{ForcLift}.

A major difference between \textsc{ForcLift} and the other two algorithms is
that \textsc{ForcLift} does not support arbitrary-precision arithmetic whereas
the other two algorithms both use the GNU Multiple Precision Arithmetic Library.
Instead, when the numerical value of the count becomes too big to be accurately
presented using finite precision, \textsc{ForcLift} returns $\infty$. We run
each algorithm on each benchmark on domains of size $2^{1}, 2^{2}, 2^{3},\dots$
until the algorithm times out after \SI{1}{\hour}, runs out of memory, or (in
the case of \textsc{ForcLift}) returns $\infty$. We measure compilation and
inference time separately, but primarily focus on total runtime (i.e., the sum
of both), which is dominated by the latter.

@@

The experiments were run on an Intel Skylake \SI{2.4}{\giga\hertz} CPU with
\SI{188}{\gibi\byte} of memory and CentOS~7 operating system. C++ programs were
compiled using the Intel C++ Compiler 2020u4. \textsc{FastWFOMC} was run on
Julia~1.10.4, while all other algorithms were run on the Java Virtual Machine
1.8.0\_201.

% 3. Results and Analysis
% Presentation of Results: Use tables, graphs, and charts to present the experimental results clearly.
% Comparison with Baselines: Compare your method’s results with the baselines.
% Discussion: Analyze and interpret the results, highlighting the strengths and potential weaknesses of your approach.
% Insights: Provide insights or patterns observed from the results.
% Challenges: Discuss any challenges or unexpected outcomes.
% Acknowledgement: Honestly acknowledge any limitations of your experiments or method.

@@

\Cref{fig:plot} summarises the experimental results. Only \textsc{FastWFOMC} and
\Cranebfs{} could lift the bijection-counting problem. On this benchmark, the
largest domain sizes that these algorithms could handle were \num{64} and
\num{4096}, respectively. On the other two benchmarks, \textsc{ForcLift} has the
lowest runtime, but, as it is finite-precision, only scales up to domain sizes
of \num{16} and \num{128} in the case of `friends' and `functions',
respectively. Next, \textsc{FastWFOMC} outperforms \textsc{ForcLift} in the case
of `friends' but not `functions': it scales up to domains of size \num{1024} and
\num{64}, respectively. Finally, both \Cranebfs{} and \Cranegreedy{} perform
nearly identically on both of these benchmarks. Similarly to the `bijections'
benchmark, \Cranetwo{} significantly surpasses the other two algorithms, scaling
up to domains of size \num{8192} and \num{67108864}, respectively.

Another aspect of the experimental results worth discussing separately is that
of compilation. As both Julia and Scala use just-in-time (JIT) compilation, both
\textsc{FastWFOMC} and \textsc{ForcLift} take longer when run on the smallest
domain size, as that is when most JIT compilation happens. As \Cranetwo{} is run
only once per benchmark, JIT compilation time is incorporated into its runtime
across all domain sizes. Moreover, while \textsc{ForcLift} compilation is
generally faster than that of \Cranetwo{}, neither is significant in comparison
to inference time. Indeed, typically, \textsc{ForcLift} compilation took around
\SI{0.5}{\second}, while \Cranetwo{} compilation took around \SI{2.3}{\second}.

% 8. Conclusion
% Summary: Briefly summarize the key findings from the experimental results.
% Future Work: Suggest directions for future research based on your findings.

So what do our experiments reveal about which algorithm should one use in
practice? If the formula can be handled by \textsc{ForcLift} and the domain
sizes are reasonably small, it is likely to be the fastest algorithm. In other
cases, \Cranetwo{} is likely to be significantly faster than \textsc{FastWFOMC}
across all domain sizes (but especially with higher domain sizes), assuming the
formula can be handled by both algorithms.

% TODO: run this section through editgpt and grammarly
